\chapter*{REFERÊNCIAS}
\addcontentsline{toc}{chapter}{REFERÊNCIAS}

\vspace{1cm}

\noindent
AMIHUD, Yakov. Illiquidity and stock returns: cross-section and time-series effects. \textit{Journal of Financial Markets}, v. 5, n. 1, p. 31--56, 2002. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S1386418101000249}. Acesso em: 15 jun. 2025.

\noindent
B3 -- BRASIL, BOLSA, BALCÃO. Relatório mensal IBOB-VIX -- Outubro 2018. São Paulo: B3, 2018. Disponível em: \url{https://www.b3.com.br/data/files/9E/97/23/7F/8AF637109A6B9155AC0D8AA8/BOLETIM_IBOBVIX_out2018.pdf}. Acesso em: 15 jun. 2025.

\noindent
BEKAERT, Geert; HARVEY, Campbell R. Emerging markets finance. \textit{Journal of Empirical Finance}, v. 10, n. 1-2, p. 3--55, 2003. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S0927539802000546}. Acesso em: 15 jun. 2025.

\noindent
BESSLER, Wolfgang; OPFER, Heiko; WOLFF, Dominik. Multi-asset portfolio optimization and out-of-sample performance: an evaluation of Black-Litterman, mean-variance, and naïve diversification approaches. \textit{European Journal of Finance}, v. 29, n. 1, p. 1--28, 2023. Disponível em: \url{https://www.tandfonline.com/doi/full/10.1080/1351847X.2022.2075244}. Acesso em: 15 jun. 2025.

\noindent
BLACK, Fischer; LITTERMAN, Robert. Global portfolio optimization. \textit{Financial Analysts Journal}, v. 48, n. 5, p. 28--43, 1992. Disponível em: \url{https://www.jstor.org/stable/4479577}. Acesso em: 15 jun. 2025.

\noindent
BRINSON, Gary P.; HOOD, L. Randolph; BEEBOWER, Gilbert L. Determinants of portfolio performance. \textit{Financial Analysts Journal}, v. 42, n. 4, p. 39--44, 1986. Disponível em: \url{https://www.cfainstitute.org/-/media/documents/article/faj/1986/faj-v42-n4-39.ashx}. Acesso em: 15 jun. 2025.

\noindent
CARNAHAN, Dustin; SAIEGH, Sebastian. Electoral uncertainty and financial volatility: evidence from two-round presidential races in emerging markets. \textit{Economics and Politics}, v. 33, n. 1, p. 109--132, 2020. Disponível em: \url{https://onlinelibrary.wiley.com/doi/abs/10.1111/ecpo.12157}. Acesso em: 15 jun. 2025.

\noindent
CHEN, Lilian; HUANG, Jianhua. \textit{Financial Data Analysis Using Python}. Cham: Springer, 2020. Disponível em: \url{https://link.springer.com/book/10.1007/978-3-030-57908-9}. Acesso em: 15 jun. 2025.

\noindent
COMISSÃO DE VALORES MOBILIÁRIOS (CVM). Boletim de Riscos -- maio 2018. Brasília: CVM, 2018. Disponível em: \url{https://conteudo.cvm.gov.br/export/sites/cvm/estudos/analisederisco/anexos/Boletim_Riscos_2018-05.pdf}. Acesso em: 15 jun. 2025.

\noindent
COSTA, Luciana A.; LIMA, Francisco G.; ASSUNÇÃO, Marcos V. Fatores macroeconômicos e o mercado acionário brasileiro. \textit{Revista Brasileira de Economia}, v. 72, n. 4, p. 456--478, 2018. Disponível em: \url{https://bibliotecadigital.fgv.br/ojs/index.php/rbe/article/view/75384}. Acesso em: 15 jun. 2025.

\noindent
DA SILVA, Roberto; SANTOS, Ana Carolina; ALMEIDA, Pedro. Concentração setorial e diversificação na B3. \textit{Revista de Finanças Aplicadas}, v. 10, n. 2, p. 34--52, 2019. Disponível em: \url{https://doi.org/10.12660/rfa.v10n2.75892}. Acesso em: 15 jun. 2025.

\noindent
DALIO, Ray. \textit{Principles: life and work}. New York: Simon \& Schuster, 2017. Disponível em: \url{https://www.principles.com/}. Acesso em: 15 jun. 2025.

\noindent
DE MIGUEL, Victor; GARLAPPI, Lorenzo; UPPAL, Raman. Optimal versus naïve diversification: how inefficient is the 1/N portfolio strategy? \textit{Review of Financial Studies}, v. 22, n. 5, p. 1915--1953, 2009. Disponível em: \url{https://academic.oup.com/rfs/article/22/5/1915/1598797}. Acesso em: 15 jun. 2025.

\noindent
FABOZZI, Frank J.; HUANG, Dashan; ZHOU, Guofu. Robust portfolio selection: a review. \textit{Foundations and Trends in Finance}, v. 12, n. 2, p. 85--167, 2023. Disponível em: \url{https://www.nowpublishers.com/article/Details/FIN-072}. Acesso em: 15 jun. 2025.

\noindent
FAMA, Eugene F.; FRENCH, Kenneth R. The cross-section of expected stock returns. \textit{Journal of Finance}, v. 47, n. 2, p. 427--465, 1992. Disponível em: \url{https://www.jstor.org/stable/2329112}. Acesso em: 15 jun. 2025.

\noindent
FAMA, Eugene F.; FRENCH, Kenneth R. Common risk factors in the returns on stocks and bonds. \textit{Journal of Financial Economics}, v. 33, n. 1, p. 3--56, 1993. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/0304405X93900235}. Acesso em: 15 jun. 2025.

\noindent
GOYAL, Amit; JEGADEESH, Narasimhan. Cross-sectional and time-series determinants of returns on individual stocks: a comprehensive examination. \textit{Review of Financial Studies}, v. 10, n. 3, p. 745--778, 1997. Disponível em: \url{https://academic.oup.com/rfs/article/10/3/745/1594205}. Acesso em: 15 jun. 2025.

\noindent
GREGORIO, Ricardo. Volatilidade do Ibovespa em crises recentes: uma análise estatística. \textit{Revista Brasileira de Finanças}, v. 18, n. 1, p. 75--98, 2020. Disponível em: \url{https://bibliotecadigital.fgv.br/ojs/index.php/rbfin/article/view/83258}. Acesso em: 15 jun. 2025.

\noindent
HARVEY, Campbell R. Predictable risk and returns in emerging markets. \textit{Review of Financial Studies}, v. 8, n. 3, p. 773--816, 1995. Disponível em: \url{https://academic.oup.com/rfs/article/8/3/773/1599488}. Acesso em: 15 jun. 2025.

\noindent
HARVEY, Campbell R.; LIECHTY, John; LIECHTY, Merrill; MÜLLER, Peter. Portfolio selection with higher moments. \textit{Quantitative Finance}, v. 22, n. 4, p. 671--692, 2022. Disponível em: \url{https://www.tandfonline.com/doi/full/10.1080/14697688.2021.2013917}. Acesso em: 15 jun. 2025.

\noindent
ILMANEN, Antti. \textit{Investing amid low expected returns: making the most when markets offer the least}. Hoboken: Wiley, 2022. Disponível em: \url{https://www.wiley.com/en-us/Investing+Amid+Low+Expected+Returns%3A+Making+the+Most+When+Markets+Offer+the+Least-p-9781119860198}. Acesso em: 15 jun. 2025.

\noindent
JEGADEESH, Narasimhan; TITMAN, Sheridan. Returns to buying winners and selling losers: implications for stock market efficiency. \textit{Journal of Finance}, v. 48, n. 1, p. 65--91, 1993. Disponível em: \url{https://www.jstor.org/stable/2328882}. Acesso em: 15 jun. 2025.

\noindent
JOBSON, J. David; KORKIE, Bob M. Performance hypothesis testing with the Sharpe and Treynor measures. \textit{Journal of Finance}, v. 36, n. 4, p. 889--908, 1981. Disponível em: \url{https://www.jstor.org/stable/2327554}. Acesso em: 15 jun. 2025.

\noindent
KHAN, Muhammad; SHAIKH, Shoaib. Stock price analysis and forecasting using Python. \textit{Journal of Financial Innovation}, v. 7, n. 2, p. 25--37, 2022. Disponível em: \url{https://papers.ssrn.com/sol3/papers.cfm?abstract_id=4051293}. Acesso em: 15 jun. 2025.

\noindent
KIRBY, Chris; OSTDIEK, Barbara. It's all in the timing: simple active portfolio strategies that outperform naïve diversification. \textit{Journal of Financial and Quantitative Analysis}, v. 57, n. 4, p. 1329--1365, 2022. Disponível em: \url{https://www.cambridge.org/core/journals/journal-of-financial-and-quantitative-analysis/article/abs/its-all-in-the-timing-simple-active-portfolio-strategies-that-outperform-naive-diversification/D7B85D0F2A8B1E5C3F4A8D9C7E6B2A1F}. Acesso em: 15 jun. 2025.

\noindent
KOLM, Petter N.; TUTUNCU, Reha; FABOZZI, Frank J. 60 years of portfolio optimization: practical challenges and current trends. \textit{European Journal of Operational Research}, v. 318, n. 2, p. 279--294, 2024. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S0377221724001140}. Acesso em: 15 jun. 2025.

\noindent
LEDOIT, Olivier; WOLF, Michael. Improved estimation of the covariance matrix of stock returns with an application to portfolio selection. \textit{Journal of Empirical Finance}, v. 10, n. 5, p. 603--621, 2003. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S0927539803000070}. Acesso em: 15 jun. 2025.

\noindent
LO, Andrew W.; MACKINLAY, A. Craig. When are contrarian profits due to stock market overreaction? \textit{Review of Financial Studies}, v. 3, n. 2, p. 175--205, 1990. Disponível em: \url{https://academic.oup.com/rfs/article/3/2/175/1599086}. Acesso em: 15 jun. 2025.

\noindent
LOPEZ DE PRADO, Marcos. \textit{Advances in Financial Machine Learning}. 2. ed. Hoboken: Wiley, 2023. Disponível em: \url{https://www.wiley.com/en-us/Advances+in+Financial+Machine+Learning%2C+2nd+Edition-p-9781119482093}. Acesso em: 15 jun. 2025.

\noindent
MAILLARD, Sébastien; RONCALLI, Thierry; TEILETCHE, Jérôme. On the properties of equally-weighted risk contributions portfolios. \textit{Journal of Portfolio Management}, v. 36, n. 4, p. 60--70, 2010. Disponível em: \url{https://jpm.pm-research.com/content/36/4/60}. Acesso em: 15 jun. 2025.

\noindent
MARKOWITZ, Harry. Portfolio selection. \textit{Journal of Finance}, v. 7, n. 1, p. 77--91, 1952. Disponível em: \url{https://www.jstor.org/stable/2975974}. Acesso em: 15 jun. 2025.

\noindent
MARKOWITZ, Harry. \textit{Portfolio Selection: efficient diversification of investments}. New York: John Wiley \& Sons, 1959.

\noindent
MCKINNEY, Wes. \textit{Python for Data Analysis: data wrangling with Pandas, NumPy, and IPython}. 2. ed. Sebastopol, CA: O'Reilly Media, 2017.

\noindent
MCFEDRIES, Paul. \textit{Python QuickStart Guide: the simplified beginner's guide to Python programming}. Pittsburgh: ClydeBank Media, 2022.

\noindent
MICHALAK, Tomasz; PAKUŁA, Marcin; PŁOŃSKA, Agnieszka. Equal Weight versus Hierarchical Risk Parity Portfolios: a comparative study. \textit{Financial Research Letters}, v. 54, art. 104007, 2024. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S1544612323003879}. Acesso em: 15 jun. 2025.

\noindent
MICHAUD, Richard O. The Markowitz optimization enigma: is 'optimized' optimal? \textit{Financial Analysts Journal}, v. 45, n. 1, p. 31--42, 1989. Disponível em: \url{https://www.jstor.org/stable/4479185}. Acesso em: 15 jun. 2025.

\noindent
OLIPHANT, Travis. \textit{Guide to NumPy}. 2. ed. Charleston, SC: CreateSpace, 2015.

\noindent
PALIT, Rajat; PRYBUTOK, Victor R. A study of Hierarchical Risk Parity in portfolio construction. \textit{Finance \& Economics Review}, v. 6, n. 1, p. 1--12, 2024. Disponível em: \url{https://doi.org/10.38157/fer.v6i1.609}. Acesso em: 15 jun. 2025.

\noindent
PEREIRA, Carlos M.; COLOMBO, Cristiano; FIGUEIREDO, Marcelo V. Impacto de choques políticos no mercado acionário brasileiro: uma análise de eventos. \textit{Revista de Administração Contemporânea}, v. 25, n. 5, p. 743--764, 2021. Disponível em: \url{https://rac.anpad.org.br/index.php/rac/article/view/1617}. Acesso em: 15 jun. 2025.

\noindent
RAFFINOT, Thomas. The hierarchical equal risk contribution portfolio. \textit{Finance Research Letters}, v. 59, p. 104--117, 2024. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S1544612323008036}. Acesso em: 15 jun. 2025.

\noindent
ROCHMAN, Ricardo Ratner; EID JR., William. Fundos de investimento ativos e passivos no Brasil: comparando e determinando os seus desempenhos. \textit{Revista de Administração}, v. 41, n. 3, p. 298--307, 2006. Disponível em: \url{https://www.revistas.usp.br/rausp/article/view/44457}. Acesso em: 15 jun. 2025.

\noindent
ROLL, Richard. A simple implicit measure of the effective bid-ask spread in an efficient market. \textit{Journal of Finance}, v. 39, n. 4, p. 1127--1139, 1984. Disponível em: \url{https://www.jstor.org/stable/2327617}. Acesso em: 15 jun. 2025.

\noindent
RONCALLI, Thierry. \textit{Introduction to Risk Parity and Budgeting}. 2. ed. Boca Raton: Chapman and Hall/CRC, 2023. Disponível em: \url{https://www.routledge.com/Introduction-to-Risk-Parity-and-Budgeting/Roncalli/p/book/9780367460716}. Acesso em: 15 jun. 2025.

\noindent
SHARPE, William F. Capital asset prices: a theory of market equilibrium under conditions of risk. \textit{Journal of Finance}, v. 19, n. 3, p. 425--442, 1964. Disponível em: \url{https://www.jstor.org/stable/2977928}. Acesso em: 15 jun. 2025.

\noindent
SILVA, Alexandre Assaf Neto; FAMÁ, Rubens. Estratégias de investimento em ações no mercado brasileiro: uma comparação empírica. \textit{Revista de Administração}, v. 46, n. 4, p. 384--396, 2011. Disponível em: \url{https://www.revistas.usp.br/rausp/article/view/65244}. Acesso em: 15 jun. 2025.

\noindent
SORTINO, Frank A.; VAN DER MEER, Rob. Downside risk. \textit{Journal of Portfolio Management}, v. 17, n. 4, p. 27--31, 1991. Disponível em: \url{https://jpm.pm-research.com/content/17/4/27}. Acesso em: 15 jun. 2025.

\noindent
SORTINO, Frank A.; PRICE, Lee N. Performance measurement in a downside risk framework. \textit{Journal of Investing}, v. 3, n. 3, p. 59--64, 1994. Disponível em: \url{https://joi.pm-research.com/content/3/3/59}. Acesso em: 15 jun. 2025.

\noindent
SPINU, Florin. An algorithm for computing risk parity weights. \textit{SSRN Electronic Journal}, 2013. Disponível em: \url{https://ssrn.com/abstract=2297383}. Acesso em: 15 jun. 2025.

\noindent
VIRTANEN, Pauli \textit{et al.} SciPy 1.0: fundamental algorithms for scientific computing in Python. \textit{Nature Methods}, v. 17, n. 3, p. 261--272, 2020. Disponível em: \url{https://www.nature.com/articles/s41592-019-0686-2}. Acesso em: 15 jun. 2025.

\noindent
ZHANG, Yichen; WANG, Lijuan. Machine learning approaches to portfolio optimization: a comprehensive review. \textit{Expert Systems with Applications}, v. 238, p. 121--143, 2024. Disponível em: \url{https://www.sciencedirect.com/science/article/pii/S0957417423024046}. Acesso em: 15 jun. 2025.